\LARGE \bf GPU Acceleration for Fixed Complexity Sphere Decoder in Large MIMO Uplink Systems

### 2015-01-05 Parallel Acceleration of Fixed Complexity Sphere Decoder Based on GPU/conf_draft/conference draft.tex
this is the final version of Parallel Acceleration of FCSD paper

% draft mode.
\documentclass[a4paper, 10pt, conference]{IEEEconf}
%\documentclass[12pt,a4paper,final]{article}
%\IEEEoverridecommandlockouts
%\overrideIEEEmargins
%\documentclass[conference]{IEEEtrans}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,ft.tex

% manually specify the path to it like:
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{caption}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{epstopdf}
% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

@@

%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{\LARGE \bf GPU Acceleration of Fixed Complexity Sphere Decoder for Large MIMO Uplink Systems}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
%\author{\IEEEauthorblockN{Tianpei Chen} and 
%\IEEEauthorblockN{Harry Leib}}
%\IEEEauthorblockA{ID: 25504946\\Department of Electrical and\\Computer Engineering\\
%McGill University\\
%Montreal, Quebec}}
\author{Tianpei Chen and Harry Leib}
 %Department of Electrical and Computer Engineering\\
 %McGill University\\
% Montreal, Quebec}

@@

\begin{document}
% make the title area
%\begin{spacing}{2.5}
\date{}
\maketitle
\thispagestyle{empty}
\pagestyle{empty}
% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Large MIMO constitutes a principal techniques for next generation wireless communication systems. Detection algorithms for large MIMO tend to require long simulation times on conventional computer systems, in order to reduce the simulation times for decoders of large MIMO systems, we propose to use General Purpose Graphic Processing Units (GPGPU). In this paper Fixed Complexity Sphere Decoder (FCSD) is implemented for the large scale MIMO system based on the dominant framework of GPGPU-Computed Unified Device Architecture (CUDA). The proposed CUDA-FCSD can achieve considerable speedup over conventional CPU implementation while reaching the same Bit Error Rate (BER) performance. Furthermore the proposed algorithm demonstrates great potential to accelerate large MIMO systems.\\
\textbf{ Keywords: Large MIMO, FCSD, GPGPU, CUDA}
\end{abstract}

@@

\section{Introduction}
The present challenge in the wireless communication area is the sharp contrast between the increasing demand of mobile network speed for the transmission of rich media content with better QoS (Quality of Service), and the shortage of radio frequency spectrum. Multiple Input Multiple Output (MIMO) is thought to be a very promising technology that can help to solve this problem owing to several advantages it can offer. The core idea of MIMO is to use multiple antennas that at the transmitting and receiving sides, that provide parallel data pipes in space resulting in spatial multiplexing gains\cite{oestges2010mimo}. For example the LTE standard supports up to 8 antennas at the base station\cite{dahlman20103g}.

Large scale MIMO (also named massive MIMO) systems employ hundreds of low-power low-price antennas collocated at the base station and serving several tens of terminals at the same time\cite{rusek2013scaling}. This technology originated from multiuser MIMO (MU-MIMO) systems\cite{larsson2013massive}. Large scale MIMO will achieve the full potential of conventional MIMO systems (spatial diversity gain and spatial multiplexing gain), as well as provide additional reduction of latency and improvement of system robustness. Detector is one of the key component of MIMO systems,
 a well known detector for MIMO system is the Sphere Decoder (SD) that perform as well as a Maximum Likelihood detector (MLD) with a dramatic complexity reduction\cite{viterbo1999universal}\cite{hassibi2005sphere}. There are two major problems with the traditional SD decoder: complexity depending on Signal to Noise Ratio (SNR), and sequential nature, which are contradictory with the requirement of a constant data processing throughput and parallel implementation for practical system. The Fixed Complexity Sphere Decoder (FCSD) overcomes these two disadvantages of the SD decoder by searching the solution along a fixed number of independent paths\cite{barbero2008fixing}.

General purpose Graphic processing Units Computing (GPGPU) has become a common trend in industry and research in areas that need massive computations, such as computer vision\cite{fung2008using}, computer games\cite{blewitt2013applicability}, signal processing\cite{van2011accelerating}\cite{6671435} and finance\cite{grauer2013accelerating}. Compared to CPU, the GPU can naturally cope with computational intensive tasks, since most of the hardware resource are allocated for data processing and parallel computation, rather than caching and flow control. In recent years GPU computing has provided significant advantages in data processing speeds over CPU computations\cite{nvidia2008programming}. In 2011, the NVIDIA GPUs have become the fastest supercomputer in the world\cite{cook2013cuda},  Computed unified Device Architecture (CUDA) is a programming model for GPGPU that was proposed by NIVIDIA and has become the dominant framework for GPGPU programming. It is based on popular programming languages(C, C++, Java, Python, Fortran, Direct Compute), such that the application developers in different areas do not need to learn a new programming language or get to know the hardware details. CUDA employs the Single Instruction Multiple Threads (SIMT) programming model, enables a GPU to implement parallel data processing, and has become a parallel computing platform for wide community of different applications.

In this paper, a GPU based implementation of Fixed Complexity Sphere Decoder is proposed, which is referred to as CUDA-FCSD. A heterogeneous programming solution for FCSD preprocessing and integrated blocked parallel path searching and decision maker solution are presented, the speedup as well as Bit Error Rate (BER) performance of different sizes of MIMO systems and signal constellations are presented and analyzed. The proposed CUDA-FCSD implementation demonstrates a great potential for large MIMO systems.

The rest of the paper is organized as follows. Section \ref{system} describes the MIMO systems model. Section \ref{programming model} introduces CUDA, GPU architecture and the programming model. In section \ref{GPUFCSD}, we present details of the implementation of CUDA-FCSD. In section \ref{simulation}, the speedup and bit error rate (BER) performance of CUDA-FCSD are presented and analyzed. Finally conclusions are drawn in section \ref{conclusion}.    
% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.

@@

\section{CUDA GPU Architecture and  Programming Model} \label{programming model}
\subsection{Memory Hierarchy}  
CUDA programming model is a C programming extension, with special defined syntax and structure. The serial code is executed on the host (CPU) side, and massive parallel data processing is performed on the device (GPU) side, this is defined as a heterogeneous programming model. A thread is the basic programming unit of CUDA, a block of cooperative threads form a thread block, in one block all the threads can share one pitch of on chip memory (shared memory), all the threads executed in one block can synchronize their action meaning the threads are permitted to execute the next instruction only when all the threads have reached the same instruction node. Blocks can be grouped into a one-dimensional, two-dimensional, or three-dimensional grid. Similarly, The threads can also be organized into one-dimensional, two-dimensional or three-dimensional blocks. The organization of the block and grid shape and dimension is determined by the application, Figure \ref{figure1}  simply illustrates the three dimensional thread hierarchy of CUDA.
\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{CUDA_programming_hierachy.eps}
\caption{CUDA 3D thread hierachy\\ $5\times 5\times 5$ grid and each block contains $5\times 5\times 5$ threads\\}
\label{figure1}
\end{figure}

\subsection{Single Instruction Multiple Thread (SIMT) Model} 
Threads are scheduled, managed, created and executed in a group of 32 threads called warp, the programs executes on the device side are defined as $\mathit{kernel}$ function, when a $\mathit{kernel}$ function is invoked, the thread blocks are partitioned into warps, the threads in one warps executes the same instruction simultaneously, the different warps are scheduled by warp scheduler, when one warp is waiting for data loading the other warps begin to execute, this is called Single Instruction Multiple Thread (SIMT) model. This programming architecture is askin to Single Instruction Multiple Data (SIMD), however unlike the SIMD vectoring machines, SIMT can branch the behavior of each threads to achieve thread-level parallelism. The GPU use massive number of threads to hide the long latency of global memory instead of using large cache as CPU. In a word CPU cores are designed to minimize latency for each thread at a time, whereas GPUs are designed to handle a large number of concurrent, lightweight threads in order to maximize throughput.
\subsection{Mircroarchitecture of GPU} 
In Figure \ref{figure2} we use the GPU-GeForce GTX 760 as an example to show the microarchitecture of  GPU. The GPU works as a coprocessor system, consisting of Stream Multiprocessors (SM). The GeForce GTX 760 has 6 SM that perform the multiply-add arithmetic operation. Furthermore, each SM has the special functional units(SFU) that can perform more complex arithmetics such as trigonometric functions and reciprocal square root. The shared memory, also called on-chip memory located on SM and has low latency and limited capacity. Global memory, which is off-chip memory of the GPU has long latency and large capacity. Another two types of the memory, texture memory and constant memory, are located off-chip, but cached. Therefore the loading speed of the data located in these two type of memory space is much faster than that from global memory. Constant memory is read only and with latency approximately as small as the registers. Registers are allocated dynamically and privately to threads, the access speed of registers is the faster than all the other four types of memories. The variables invoked in $\mathit{kernel}$ functions can be stored in either of the these five types of memories.

Because of the resource limitation of SM the total number of threads can be defined in one block is limited. On current CUDA supported GPUs, one block can contains up to 1024 threads, and all the blocks are executed on SM. Thus the total number of threads ($T_{total}$) we can make use of is $T_{total}=N_{SM}\times N_{threads/SM} $, $N_{SM}$ and $N_{threads/SM}$ denote the number of SM and number of threads per SM. For example the GeForce GTX 760 allows 2048 threads in one SM, thus the total number of threads that can work in parallel is $2048\times 6=12288$.
\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{High_view_of_CUDA_GPU_microarchitecture.eps}
\caption{High View of CUDA CPU Microarchitecture}
\label{figure2}
\end{figure}   
% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.
\section{GPU Based Acceleration of FCSD}\label{GPUFCSD}
\subsection{MIMO System Model}\label{system}
% no \IEEEPARstart
%This demo file is intended to serve as a ``starter file''
%for IEEE conference papers produced under \LaTeX\ using
%IEEEtran.cls version 1.8 and later.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
%I wish you the best of success.
%\hfill mds
%\hfill December 27, 2012

We consider a complex uncoded spatial multiplexing MIMO system with $N_r$ receive and $N_t$ transmit antennas, $N_{r}\geq N_{t}$, over a flat fading channel, corresponding to the discrete time model, the $\mathbb{y}\in\mathbb{C}^{N_{r}\times 1}$ received symbol vecot can be written as:
\begin{equation}
\mathbf{y}=\mathbf{H}\mathbf{s}+\mathbf{n}   \label{formula 1}
\end{equation}
$\mathbf{s}\in \mathbb{C}^{N_{t}\times 1}$ is the transmitted symbol vector, with components that are mutually independent and taken from the signal constellation $\mathbb{O}$ (4-QAM, 16-QAM, 64-QAM), of size $M$. All the possible transmitted symbol vectors forms a $N_{t}$ complex signal vector space $\mathbb{O}^{N_{t}}$. $\mathbb{E}[\mathbf{s}\mathbf{s}^{H}]=\mathbf{I}_{N_t}E_{s}$, where $E_{s}$ denotes the symbol average energy, and $\mathbb{E}[\cdot]$ denotes the expectation operation. $\mathbf{H}\in \mathbb{C}^{N_{r}\times N_{t}}$ denotes the Rayleigh fading channel propagation matrix with independent identity distributed(i.i.d) circularly symmetric complex Gaussian zero mean components of unit variance, $\mathbf{n}\in \mathbb{C}^{N_{r}\times 1}$ is the additive white Gaussian noise (AWGN) samples with zero mean components and $\mathbb{E}[\mathbf{n}\mathbf{n}^{H}]=\mathbf{I}_{N_{r}}N_{0}$, $N_{0}$ denotes the average noise power spectrum density and $\frac{E_{s}}{N_{0}}$ denotes the signal to noise ratio (SNR).

Assume we have the perfect channel state information (CSI), meaning that $ \mathbf{H}$ is known at the receiver, as well as the SNR. The task of MIMO decoder is to recover $\mathbf{s}$ based on $\mathbf{y}$ and $\mathbf{H}$.
\subsection{Fixed Complexity Sphere Decoder}
From (\ref{formula 1}), the maximum likelihood detector (MLD) for a MIMO system is specified by:
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{y}-\mathbf{H}\mathbf{s}||^{2} \label{formula 2}
\end{equation}
We consider a MIMO system with  $N_{t}=N_{r}$. Performing the QR decomposition to the channel propagation matrix
\begin{equation}
 \mathbf{H}=\mathbf{Q}\mathbf{R}  \label{QR}
\end{equation}
where $\mathbf{Q}\in \mathbf{C}^{N_{t}\times N_{t}}$ is an unitary matrix and $\mathbf{R}\in \mathbf{C}^{N_{t}\times N_{t}}$, is an upper triangular matrix, the basing on \cite{golub2012matrix}, we can write (\ref{formula 2}) as:
\begin{eqnarray}
\nonumber
\mathbf{s}_{ML}&=&arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{Q}^{H}\mathbf{H}\mathbf{s}||^{2}\\
&=& arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{R}\mathbf{s}||^{2} \label{formula 3}
\end{eqnarray}
Consider the unconstrained estimation of $\mathbf{s}$, $\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}$, $\mathbf{G}=(\mathbf{H}^{H}\mathbf{H})^{-1}\mathbf{H}^{H}$ based on (\ref{QR}), we have  
\begin{equation}
\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}=(\mathbf{R}^{H}\mathbf{R})^{-1}\mathbf{R}^{H}\mathbf{Q}^{H}\mathbf{y}
=\mathbf{R}^{-1}\mathbf{Q}^{H}\mathbf{y}
\label{unconstrained estimation}
\end{equation} 
Therefore $\mathbf{Q}^{H}\mathbf{y}=\mathbf{R}\mathbf{\hat{s}}$, and (\ref{formula 3}) can be written as
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{R}(\mathbf{\hat{s}}-\mathbf{s})||^{2} \label{formula 4}
\end{equation}

Just like SD the FCSD also performs depth first searching, at the first $\rho$ node levels (\cite{barbero2008fixing} shows when $N_{r}=N_{t}$ the FCSD can achieve the same diversity as the MLD if $\rho\geq \sqrt{N_{t}}-1$) the FCSD search all the possible signal symbols in constellation $\mathbb{O}^{N_{t}}$, which is called Full Expansion(FE) stage, therefore for the symbol nodes $\widetilde{s}_{N_{t}} \widetilde{s}_{N_{t}-1}\dots \widetilde{s}_{N_{t}-\rho+1}$, the FCSD has $M$ branches expansion at each symbol node,  as shown in Figure \ref{figure4}, there are $\rho=2$ FE node levels and each node level has $M=4$ branches. According to (\ref{formula 4}), we have: 
\begin{equation}
\hat{s}^{F}_{i}=arg\min_{s\in \mathbb{O}^{N_{t}}}\{\sum_{i=1}^{N_{t}}|r_{ii}(\hat{s}_{i}-s_{i})+\sum_{j=i+1}^{N_{t}}r_{ij}(\hat{s}_{j}-s_{j})|^{2}\}  \label{formula 5}
\end{equation} 
$i\in [1,2\dots N_{t}]$, where $r_{ij}$ denotes the component at the $i$th row and $j$th column of the upper triangular matrix $\mathbf{R}$,  based on (\ref{formula 5}) at the remaining $N_{t}-\rho$ levels of symbols, the FCSD performs decision feed back, therefore there are only $1$ branch expansion for each symbol node at this stage, this stage is called Single Expansion (SE). 
Thus the total number of the branches is fixed (For example in Figure \ref{figure4} the total branch number is $4^{2}$), the FCSD works as a constant number of multiple path tree searching algorithm.
\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{FCSD_tree_searching.eps}
\caption{Tree searching of 4-QAM FCSD for 8x8 MIMO system}
\label{figure4}
\end{figure}
In conclusion the FCSD symbol vector candidate $s^{F}_{i}=[s_{1},s_{2}\dots s_{N_{t}}]$ can be expressed as:
\begin{displaymath}
s^{F}_{i}=\left\lbrace\begin{array}{c}

>>>>

% draft mode.
%\documentclass[a4paper, 10pt, conference]{ieeeconf}
\documentclass[letterpaper, 10pt, conference]{ieeeconf}
%\documentclass[12pt,a4paper,final]{article}
%\IEEEoverridecommandlockouts
%\overrideIEEEmargins
%\documentclass[conference]{IEEEtrans}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,ft.tex

% manually specify the path to it like:
\usepackage{amssymb}
%\usepackage{amsthm}
%\usepackage{caption}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{epstopdf}
%\usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
%\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{fancyhdr}
% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

@@

%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{\LARGE \bf GPU Acceleration for Fixed Complexity Sphere Decoder in Large MIMO Uplink Systems}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
%\author{\IEEEauthorblockN{Tianpei Chen} and 
%\IEEEauthorblockN{Harry Leib}}
%\IEEEauthorblockA{ID: 25504946\\Department of Electrical and\\Computer Engineering\\
%McGill University\\
%Montreal, Quebec}}
\author{Tianpei Chen and Harry Leib\\
Department of Electrical and Computer Engineering, McGill University, Montreal, Quebec\\
tianpei.chen@mail.mcgill.ca harry.leib@mcgill.ca}
 %Department of Electrical and Computer Engineering\\
 %McGill University\\
% Montreal, Quebec}

@@

\begin{document}
% make the title area
%\begin{spacing}{2.5}
%\date{}
\maketitle
\thispagestyle{fancy} % IEEE模板在\maketitle后会自动声明\thispagestyle{plain}，  
                      % 导致第一页什么都没有。所以得把plain更改为fancy  
\lhead{} % 页眉左，需要东西的话就在{}内添加  
\chead{} % 页眉中  
\rhead{\thepage} % 页眉右  
\lfoot{} % 页眉左  
\cfoot{} % 页眉中  
\rfoot{} %页眉右，\thepage 表示当前页码  
\renewcommand{\headrulewidth}{0pt} %改为0pt即可去掉页眉下面的横线  
\renewcommand{\footrulewidth}{0pt} %改为0pt即可去掉页脚上面的横线  
\pagestyle{fancy}  
\rhead{\thepage}
%\thispagestyle{empty}
%\pagestyle{empty}
% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Large MIMO constitutes a principal technology for next generation wireless communication systems. Detection algorithms for large MIMO schemes tend to require long simulation times on conventional computer systems. In order to improve simulation efficiency of decoders in large MIMO systems, we propose to use General Purpose Graphic Processing Units (GPGPU). This paper considers the implementation of a Fixed Complexity Sphere Decoder (FCSD) for large scale MIMO systems using the framework of GPGPU-Computed Unified Device Architecture (CUDA). The proposed CUDA-FCSD can achieve considerable speedups over conventional CPU implementation while providing the same Bit Error Rate (BER) performance. Furthermore the proposed algorithm demonstrates the great potential of GPGPU to accelerate the simulation of large MIMO systems.

\textbf{ Keywords: Large MIMO, fixed complexity sphere decoder (FCSD), general purpose graphic processing units (GPGPU), computed unified device architecture (CUDA)}
\end{abstract}

@@

\section{Introduction}
A present challenge in the wireless communication area is to bridge the sharp contrast between the increasing demand of mobile network speed for the transmission of rich media content with better QoS (Quality of Service), and the shortage of radio frequency spectrum. Multiple Input Multiple Output (MIMO) technology is of immense research interest since it can help solving this problem owing to several advantages it offers. The core idea of MIMO is to use multiple antennas at the transmitting and receiving sides, that provide parallel data pipes, resulting in spatial multiplexing and diversity gains\cite{oestges2010mimo}.

Large scale MIMO (also named massive MIMO) systems employ tens to hundreds of low-power low-price antennas at the base station, that serve many terminals at the same time\cite{rusek2013scaling}. Large scale MIMO can achieve the full potential of conventional MIMO systems, as well as provide additional reduction of latency and improvement of system robustness. The detector is one of the key components of MIMO systems.
 A well known detector for MIMO systems is the Sphere Decoder (SD) that performs as well as a Maximum Likelihood detector (MLD), with a dramatic complexity reduction\cite{viterbo1999universal}\cite{hassibi2005sphere}. There are two major problems with the traditional SD decoder: complexity depending on Signal to Noise Ratio (SNR), and sequential nature, which conflict with the requirement of a constant data processing throughput and parallel implementation for practical systems. The Fixed Complexity Sphere Decoder (FCSD) overcomes these problems of the SD decoder by searching the solution along a fixed number of independent paths\cite{barbero2008fixing}.

General Purpose Graphic Processing Units (GPGPU) computing is a common trend in industry and research in areas that need massive computations, such as computer vision\cite{fung2008using}, computer games\cite{blewitt2013applicability}, signal processing\cite{van2011accelerating}\cite{6671435}, and finance\cite{grauer2013accelerating}. Compared to a CPU, a GPU can naturally cope with computational intensive tasks, since most of the hardware resources are allocated for data processing and parallel computation, rather than caching and flow control\cite{cook2013cuda}. In recent years GPU computing has provided significant advantages in data processing speeds over CPU computations\cite{nvidia2008programming}. The Computed unified Device Architecture (CUDA)\cite{cook2013cuda}\cite{nvidia2008programming} is a programming model for GPGPU that was proposed by NIVIDIA and has become the dominant framework for GPGPU programming. It is based on popular programming languages (C, C++, Java, Python, Fortran, Direct Compute), such that the application developers in different areas do not need to learn a new programming language or get to know the hardware details. CUDA employs the Single Instruction Multiple Threads (SIMT) programming model, enables a GPU to implement parallel data processing, and has become a parallel computing platform for different applications.

In this paper, a GPU based implementation of Fixed Complexity Sphere Decoder (FCSD) is proposed, which is referred to as CUDA-FCSD. A heterogeneous programming solution for FCSD preprocessing and integrated blocked parallel path searching and decision making are presented. Using this implementation the speedup as well as Bit Error Rate (BER) performance of different sizes of MIMO systems and signal constellations are analyzed. The proposed CUDA-FCSD implementation demonstrates a great potential for large MIMO systems.

The rest of the paper is organized as follows. Section \ref{programming model} briefly introduces CUDA, the GPU architecture and the programming model. In section \ref{GPUFCSD}, we present details of the implementation of CUDA-FCSD. In section \ref{simulation}, the speedup and bit error rate (BER) performance of CUDA-FCSD are considered. Finally, conclusions are drawn in section \ref{conclusion}.    
% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.

@@

\section{CUDA Programming Model and GPU Architecture} \label{programming model} 
\subsection{Single Instruction Multiple Thread Model}
The CUDA programming model is a C programming extension with special defined syntax and structure. The serial code is executed on the host (CPU) side, and massive parallel data processing is performed on the device (GPU) side. Such an approach is defined as a heterogeneous programming model. A thread is the basic programming unit of CUDA. A block of cooperative threads form a thread block, where they can synchronize their behavior and communicate. Fig. \ref{figure1} illustrates a simple CUDA memory hierachy model.
\begin{figure}[htb]
\centering
\includegraphics*[width=0.5\textwidth]{CUDA_programming_hierachy.eps}
\caption{ CUDA 3D memory hierachy $5\times 5\times 5$ grid, with each block composed of $5\times 5\times 5$ threads.
The threads can be organized into one-dimensional, two-dimensional or three-dimensional blocks. Similarly, blocks can also be grouped into a one-dimensional, two-dimensional, or three-dimensional grid.
}
\label{figure1}
\end{figure}

Threads are scheduled, managed, created and executed in a group of 32, called warp. The programs that execute on the device side are defined as \textit{kernel} functions. When a \textit{kernel} function is invoked, the thread blocks are partitioned into warps. The threads in one warp execute the same instruction simultaneously. The different warps are scheduled by a warp scheduler. When one warp is waiting for data loading, the other warps begin to execute. Such an approach is called a Single Instruction Multiple Thread (SIMT) model. This programming architecture is akin to a Single Instruction Multiple Data (SIMD) scheme, however unlike SIMD vectoring machines, the SIMT can branch the behavior of each threads to achieve thread-level parallelism. The GPU uses a massive number of threads to hide the long latency of a global memory, instead of using a large cache as CPU. A CPU core is designed to minimize the latency for each thread at a time, whereas a GPU is designed to handle a large number of concurrent lightweight threads, in order to maximize throughput.
\subsection{Mircroarchitecture of GPU} 
We present the GPU-GeForce GTX 760 microarchitecture in Fig. \ref{figure2} as an example. The GPU works as a coprocessor system, consisting of Stream Multiprocessors (SMs), which perform multiply-add arithmetic operations. Furthermore, each SM has the special functional units (SFU) that can perform more complex operations such as trigonometric functions and reciprocal square root. There are five types of memories: registers, shared memory, global memory, constant memory and texture memory. Registers are allocated dynamically and privately to threads, and provide most rapid access speeds. The shared memory, also called on-chip memory, is located on the SMs and can be shared among the threads in one block. Shared memory has low latency and limited capacity. Global memory, which is off-chip has large latency and large capacity. The other two types of memories, constant memory and texture memory, are located off-chip, but cached. Therefore the access speed of these two type of memory space is much faster than that of global memory. The variables invoked by \textit{kernel} functions can be stored in either of the these five types of memories.

Because of the resource limitation of SM, the total number of threads that can be defined in one block is limited. On current CUDA supported GPUs, one block contains up to 1024 threads. Thus the total number of threads ($T_{total}$) we can use is $T_{total}=N_{SM}\times N_{threads/SM} $, where $N_{SM}$ and $N_{threads/SM}$ denote the number of SMs and number of threads per SM.  
\begin{figure}[htb]
\centering
\includegraphics[width=0.45\textwidth]{High_view_of_CUDA_GPU_microarchitecture.eps}
\caption{ High View of CUDA CPU Microarchitecture
The GeForce GTX 760 has 6 SMs and the maximum number of threads allowed on each SM is 2048. Thus the total number of threads that can work in parallel is $2048\times 6=12288$.
}
\label{figure2}
\end{figure}   
% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.
\section{GPU Based Acceleration of FCSD}\label{GPUFCSD}
\subsection{MIMO System Model}\label{system}
% no \IEEEPARstart
%This demo file is intended to serve as a ``starter file''
%for IEEE conference papers produced under \LaTeX\ using
%IEEEtran.cls version 1.8 and later.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
%I wish you the best of success.
%\hfill mds
%\hfill December 27, 2012

We consider a complex uncoded spatial multiplexing MIMO system with $N_r$ receive and $N_t$ transmit antennas, $N_{r}\geq N_{t}$, over a flat fading channel. Using a discrete time model, $\mathbf{y}\in\mathbb{C}^{N_{r}\times 1}$ is the received symbol vector written as:
\begin{equation}
\mathbf{y}=\mathbf{H}\mathbf{s}+\mathbf{n},   \label{formula 1}
\end{equation}
where $\mathbf{s}\in \mathbb{C}^{N_{t}\times 1}$ is the transmitted symbol vector, with components that are mutually independent and taken from a signal constellation $\mathbb{O}$ (4-QAM, 16-QAM, 64-QAM) of size $M$. All the possible transmitted symbol vectors $\mathbf{s}\in \mathbb{O}^{N_{t}}$, with $\mathbb{E}[\mathbf{s}\mathbf{s}^{H}]=\mathbf{I}_{N_t}E_{s}$, where $E_{s}$ denotes the symbol average energy, and $\mathbb{E}[\cdot]$ denotes the expectation operation. Furthermore $\mathbf{H}\in \mathbb{C}^{N_{r}\times N_{t}}$ denotes the Rayleigh fading channel propagation matrix with independent identity distributed (i.i.d) circularly symmetric complex Gaussian components of zero mean and unit variance. Finally, $\mathbf{n}\in \mathbb{C}^{N_{r}\times 1}$ is the additive white Gaussian noise (AWGN) vector with zero mean components and $\mathbb{E}[\mathbf{n}\mathbf{n}^{H}]=\mathbf{I}_{N_{r}}N_{0}$, where $N_{0}$ denotes the noise power spectrum density, and hence $\frac{E_{s}}{N_{0}}$ is the signal to noise ratio (SNR).

Assume the receiver has the perfect channel state information (CSI), meaning that $ \mathbf{H}$ is known by receiver, as well as the SNR. The task of the MIMO decoder is to recover $\mathbf{s}$ based on $\mathbf{y}$ and $\mathbf{H}$.
\subsection{Fixed Complexity Sphere Decoder}
From (\ref{formula 1}), the maximum likelihood detector (MLD) for a MIMO system is specified by:
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{y}-\mathbf{H}\mathbf{s}||^{2}. \label{formula 2}
\end{equation}
We consider a MIMO system with  $N_{t}=N_{r}$. Performing the QR decomposition of the channel propagation matrix
\begin{equation}
 \mathbf{H}=\mathbf{Q}\mathbf{R},  \label{QR}
\end{equation}
where $\mathbf{Q}\in \mathbb{C}^{N_{t}\times N_{t}}$ is a unitary matrix and $\mathbf{R}\in \mathbb{C}^{N_{t}\times N_{t}}$, is an upper triangular matrix \cite{golub2012matrix}, we can write (\ref{formula 2}) as:
\begin{eqnarray}
\nonumber
\mathbf{s}_{ML}&=&arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{Q}^{H}\mathbf{H}\mathbf{s}||^{2}\\
&=& arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{R}\mathbf{s}||^{2}. \label{formula 3}
\end{eqnarray}
Using the unconstrained estimator of $\mathbf{s}$, $\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}$, $\mathbf{G}=(\mathbf{H}^{H}\mathbf{H})^{-1}\mathbf{H}^{H}$, and (\ref{QR}), we have  
\begin{equation}
\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}=(\mathbf{R}^{H}\mathbf{R})^{-1}\mathbf{R}^{H}\mathbf{Q}^{H}\mathbf{y}
=\mathbf{R}^{-1}\mathbf{Q}^{H}\mathbf{y}.
\label{unconstrained estimation}
\end{equation} 
Therefore $\mathbf{Q}^{H}\mathbf{y}=\mathbf{R}\mathbf{\hat{s}}$, and (\ref{formula 3}) can be written as
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{R}(\mathbf{\hat{s}}-\mathbf{s})||^{2}. \label{formula 4}
\end{equation}

Similarly to SD the FCSD also performs depth first searching. At the first $\rho$ node levels\footnote{when $N_{r}=N_{t}$ the FCSD can achieve the same diversity as the MLD if $\rho\geq \sqrt{N_{t}}-1$ \cite{barbero2008fixing}} the FCSD searches all the possible signal symbols in constellation $\mathbb{O}^{N_{t}}$, a process called the Full Expansion (FE) stage. Let $\mathbf{s}^{F}$ denote FCSD symbol vector candidates. For the symbols $s^{F}_{N_{t}}, s^{F}_{N_{t}-1}\dots s^{F}_{N_{t}-\rho+1}$, the FCSD has $M$ branch expansions at each symbol node. As shown in Fig. \ref{figure4}, there are $\rho=2$ FE node levels and each node level has $M=4$ branches. According to (\ref{formula 4}), we have: 
\begin{equation}
s^{F}_{i}=arg\min_{s\in \mathbb{O}^{N_{t}}}\{\sum_{i=1}^{N_{t}}|r_{ii}(\hat{s}_{i}-s^{F}_{i})+\sum_{j=i+1}^{N_{t}}r_{ij}(\hat{s}_{j}-s^{F}_{j})|^{2}\},  \label{formula 5}
\end{equation} 
$i\in [1,2\dots N_{t}]$, where $r_{ij}$ denotes the component at the $i$th row and $j$th column of the upper triangular matrix $\mathbf{R}$. Based on (\ref{formula 5}) at the remaining $N_{t}-\rho$ levels of symbols, the FCSD employs decision feedback, and therefore there are only one branch expansion for each symbol node at this stage. This stage is called Single Expansion (SE). 
Thus the total number of the branches is fixed, and the FCSD works as a constant number of multiple path tree searching algorithm.
